Staged Abstract Interpreters

### 2019-08-14 paper/10_intro.tex
revision

\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
  \item We begin by introducing our target language and reviewing
    monads in Scala, and then presenting the generic interpreter (Section~\ref{prelim}).
    After which, we review its instantiations of concrete interpretation
    (Section~\ref{unstaged_conc}) and staged concrete interpretation
    (Section~\ref{stagedinterp}).
  \item We present the unstaged abstract interpreter under the same framework by
    replacing the environment, store and values to their abstract counterparts (Section~\ref{unstaged_abs}).
    After, we show the combination of approximation and specialization, dubbed
    \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
  \item We conduct three case studies (Section~\ref{cases_study}), which demonstrate that
    our approach requires less engineering efforts, is applicable to various analysis,
    and enables compiling analysis modularly.
  \item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

>>>>

\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
  \item We begin by introducing our target language and reviewing
    monads in Scala, and then presenting the generic interpreter
    (Section~\ref{prelim}).  After which, we review its instantiations
    of concrete interpretation (Section~\ref{unstaged_conc}) and
    staged concrete interpretation (Section~\ref{stagedinterp}).
  \item We present the unstaged abstract interpreter under the same
    framework by replacing the environment, store and values to their
    abstract counterparts (Section~\ref{unstaged_abs}). Then we
    show the combination of approximation and specialization, dubbed
    \textit{staged abstract interpreters}, can be readily derived
    (Section~\ref{sai}). We also summarize the approach and discuss
    the soundness issue after showing the four artifacts.
  \item We conduct three case studies (Section~\ref{cases_study}) and
    an empirical evaluation on performance improved (Section~\ref{evaluation}).
    Finally we discuss related work and conclude.
\end{itemize}

### 2019-08-14 paper/20_prelim.tex
revision

In this section, we first describe the abstract syntax of the language for our interpreter, 
then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monad to demonstrate the
idea, but the approach is not restricted to our choice. One can use
imperative or direct style in other MSP languages (e.g., MetaOCaml
\cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} and Template
Haskell \cite{Sheard:2002:TMH:636517.636528} to construct such staged abstract
interpreters.

@@

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore it only exposes the simplified version of @flatMap@, where the
monad @M[A]@ is not introduced as a function argument. The trait @MonadOps@
(Figure \ref{fig:monadops}) defines the simplified version of monadic
operations that are necessary for @for@-comprehension. 
The conversion between @Monad@ and @MonadOps@ can be done by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and monad
transformers such as @ReaderT@, @StateT@, and @ListT@ to write our interpreters.
The implementation of monads and monad transformers essentially borrows the
ground-truth from Haskell.

@@

Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
take a monad as argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one. Constructing
extensible interpreters using monad transformers was first proposed by
\citet{DBLP:conf/popl/LiangHJ95}, and later has been applied to abstract
interpreters \cite{Sergey:2013:MAI:2491956.2491979,
DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}. In this section,
with the multi-stage programming and monad transformers in mind while leaving them
as abstract type members, we present the generic interface of a big-step
definitional interpreter.

\paragraph{Basic Types} We start with some basic type definitions used in the
interpreter. The identifiers in the program are represented by strings. The two
default components in the interpreter are environments @Env@ and stores @Store@,
i.e., mappings from identifiers to addresses and mappings from addresses to
values, respectively. The @Env@ models accessible variables, and @Store@ models
the persistent heap through the program runtime. But the @Addr@ and @Value@ are
just declared as abstract types.

@@

\paragraph{Binding-time Abstraction} As mentioned before, the binding-time is
declared as a higher-kinded type @R[_]@. If we simply instantiate @R@ as an identity
type (i.e., @type R[T] = T@), then the generic interpreter will execute the program.
In Section \ref{stagedinterp}, we will instantiate @R@ using LMS's built-in
next-stage type annotation @Rep@, which makes the interpreter act as a compiler.

\paragraph{Monadic Operations} We define the return type of the interpreter as
@Ans@, which is an arbitrary monad type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain constraints have to be added on the type @AnsM@. Here, we use a
structural type @MonadOps@ to require @AnsM@ to at least implement @map@ and
@flatMap@. It is worth noting that @MonadOps@ takes another type parameter
@R[_]@ as binding-time; accordingly the outer @R@ defined in the trait is passed
to @MonadOps@. Inside of @MonadOps@, @R[_]@ wraps the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When acting as
compilers, we will also replace the monads to the ones that work on staged
values.

We also define several methods to obtain and modify the environment and store.
These methods return monadic values of type @AnsM[_]@, which encapsulate the 
environment or store, or simply a @Unit@ value for effects.
For example, @local_env@ takes a value @ans@ of type @Ans@ and an environment 
which will be installed when evaluating @ans@.

@@

\paragraph{Primitive Operations} Next we define several primitive operations.
Two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is also wrapped
by @R[_]@. The other one would simply wrap the address with our monadic type
@AnsM[_]@.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

Other primitive operations provide basic functionality for the interpreter.
The method @num@ and @close@ deal with primitive values, which lift literal
terms (e.g., lambdas) to our value representation (e.g., closures).
The method @get@ simply retrieves the value mapped from the identifier @x@ in
the environment and store. Conditionals and arithmetic is handled by @br0@
and @arith@, respectively. The methods @ap_clo@ takes a function value and an
argument value and then does the application. Note that the @Env@, @Store@, and
@Value@ are all wrapped by @R[_]@ since they will be known as next-stage value
when acting as compilers.

%TODO: close/ap_clo takes an ev

@@

\paragraph{The Interpreter} Now we can define the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}.
The essential idea is to traverse the abstract syntax tree while maintaining the
effects such as reader and state.
It is worth noting that the interpreter is written in open-recursive style -- it
can not refer to itself directly, instead, @eval@ takes an additional parameter
@ev@ of type @Expr => Ans@ to refer to itself. Accordingly, the method @close@
that lifts lambda terms to closures and @ap_clo@ that applies functions also
takes an extra @ev@.

%\vspace{-1em}
\begin{figure}[h!]
  \centering
  \begin{lstlisting}
          def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
            case Lit(i) => num(i)                   case Let(x, rhs, e) => for {
            case Var(x) => for {                      v  <- ev(rhs)
              ρ <- ask_env                            ρ  <- ask_env
              σ <- get_store                          α  <- alloc(x)
            } yield get(σ, ρ, x)                      _  <- set_store(α → v)
            case Lam(x, e) => for {                   rt <- local_env(ev(e))(ρ + (x → α))
              ρ <- ask_env                          } yield rt
            } yield close(ev)(Lam(x, e), ρ)         case Aop(op, e1, e2) => for {
            case App(e1, e2) => for {                 v1 <- ev(e1)                                               
              v1 <- ev(e1)                            v2 <- ev(e2)
              v2 <- ev(e2)                          } yield arith(op, v1, v2)
              rt <- ap_clo(ev)(v1, v2)              case Rec(x, rhs, e) => for {
            } yield rt                                α  <- alloc(x)
            case If0(e1, e2, e3) => for {             ρ  <- ask_env
              cnd <- ev(e1)                           v  <- local_env(ev(rhs))(ρ + (x → α))
              rt  <- br0(cnd, ev(e2), ev(e3))         _  <- set_store(α → v)
            } yield rt                                rt <- local_env(ev(e))(ρ + (x → α))
                                                    } yield rt                    
          }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}
%\vspace{-1em}

Since the interpreter is written in an open-recursive style, we declare an abstract
combinator @fix@ used to close the recursion. For concrete-interpretation instantiation, it
works like the Y combinator; for abstract-interpretation instantiation, we will
instrument the interpreter by defining a memorized version of @fix@ to ensure
termination. Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also an abstract type.

>>>>

In this section, we first describe the abstract syntax of the language for our interpreters. 
Then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one.
It is worth noting that we choose to use Scala and monadic style to
demonstrate the idea, but the approach is not restricted to our
choice. One can use imperative or direct style in other MSP languages
(e.g., MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14}
and Template Haskell \cite{Sheard:2002:TMH:636517.636528} to construct
such staged abstract interpreters.

@@

Note that here the monadic object @List[_]@ encapsulates the data
internally.  Therefore it only exposes the simplified version of
@flatMap@, where the monad @M[A]@ is not introduced as a function
argument. The trait @MonadOps@ (Figure \ref{fig:monadops}) defines the
simplified version of monadic operations that are necessary for
@for@-comprehension. The conversion between @Monad@ and @MonadOps@
can be done by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax
and a few of monads and monad transformers such as @ReaderT@,
@StateT@, and @ListT@ to write our interpreters. The implementation
of monads and monad transformers essentially borrows the ground-truth
from Haskell.

@@

Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
take a monad as argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one. Constructing
extensible interpreters using monad transformers was first explored by
\citet{DBLP:conf/popl/LiangHJ95}, and later has been applied to abstract
interpreters \cite{Sergey:2013:MAI:2491956.2491979,
  DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
In this section, we present the generic inferface in the style of big-step
definitional interpreter. The key idea is to keep the binding-time type and
returned monadic type abstract, so that they can be instantiated differently.
We also need to abstract the primitive operations on those types.

\paragraph{Basic Types} We start with some basic type definitions used in the
interpreter. The identifiers in the program are represented by
strings. To represent states, two required components for the
interpreter are environments @Env@ and stores @Store@. @Env@ maps from
identifiers to addresses, and @Store@ maps from addresses to values,
respectively. The @Env@ captures bound variables in scope, and @Store@
models a persistent heap through the program runtime.  At this moment,
the domains of addresses and values are still abstract, i.e., they are
declared as abstract types.

@@

\paragraph{Binding-time Abstraction} As mentioned before, the binding-time is
declared as a higher-kinded type @R[_]@. We are free to use @R@ to annotate
other datatypes in the interpreter; it is also injected into the monadic
type @MonadOps@.
If we simply instantiate @R@ an identity type (i.e., @type R[T] = T@),
then the generic interpreter is a standard definitional interpreter
that will execute the program.  In Section \ref{stagedinterp}, we will
instantiate @R@ using LMS's built-in next-stage type annotation @Rep@,
which makes the interpreter act as a compiler.

\paragraph{Monadic Operations} We define the return type of the interpreter as
@Ans@, which is a monad type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain operations have to be added on the type @AnsM@. Here, we use a
structural type @MonadOps@ to require that type @AnsM@ at least implements @map@ and
@flatMap@. It is worth noting that @MonadOps@ takes another type parameter
@R[_]@ as binding-time that annotates data; accordingly the type @R@
defined in the outer scope is passed to @MonadOps@. Inside of @MonadOps@,
@R[_]@ wraps the data types @A@ and @B@ that are encapsulated by the
monad, but not the monad type @M@ itself. When acting as compilers, we
will also replace the monads to the ones that work on staged data values.
\todo{it is important that R only annotates data, not the monad}

We also declare several operations to mannipulate environments and
stores. These methods return monadic values of type @AnsM[_]@, which
may encapsulate over the environment or store, or simply a @Unit@
value for effects. For examples, @local_env@ installs a new environment
@ρ@ when evaluating the monadic value @ans@; @set_store@ takes a pair of
addresses and values (potentially staged) and updates the store accordingly.

@@

\paragraph{Primitive Operations} Next we define several primitive operations.
Two versions of @alloc@ are declared. The first one takes a store and an
identifier and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is therefore
wrapped by @R[_]@. The other @alloc@ is simply the monadic version of the addresses.

\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

Other primitive operations provide basic functionality for the
interpreter.  The method @num@ and @close@ deal with primitive values,
which lift literal terms (e.g., lambdas) to our value representation
(e.g., closures).  The method @get@ retrieves the value of an
identifier @x@, through the environment and store. Conditionals and
arithmetic is handled by @br0@ and @arith@, respectively. The method
@ap_clo@ is used for function applications: it takes a function value
and an argument value. Note that the @Env@, @Store@, and @Value@ are
all annotated by @R[_]@ since they can potentially be next-stage
values when acting as compilers.

@@

\begin{figure}[h!]
  \centering
  \begin{lstlisting}
          def eval(ev: Expr => Ans)(e: Expr): Ans = e match {
            case Lit(i) => num(i)                   case Let(x, rhs, e) => for {
            case Var(x) => for {                      v  <- ev(rhs)
              ρ <- ask_env                            ρ  <- ask_env
              σ <- get_store                          α  <- alloc(x)
            } yield get(σ, ρ, x)                      _  <- set_store(α → v)
            case Lam(x, e) => for {                   rt <- local_env(ev(e))(ρ + (x → α))
              ρ <- ask_env                          } yield rt
            } yield close(ev)(Lam(x, e), ρ)         case Aop(op, e1, e2) => for {
            case App(e1, e2) => for {                 v1 <- ev(e1)                                               
              v1 <- ev(e1)                            v2 <- ev(e2)
              v2 <- ev(e2)                          } yield arith(op, v1, v2)
              rt <- ap_clo(ev)(v1, v2)              case Rec(x, rhs, e) => for {
            } yield rt                                α  <- alloc(x)
            case If0(e1, e2, e3) => for {             ρ  <- ask_env
              cnd <- ev(e1)                           v  <- local_env(ev(rhs))(ρ + (x → α))
              rt  <- br0(cnd, ev(e2), ev(e3))         _  <- set_store(α → v)
            } yield rt                                rt <- local_env(ev(e))(ρ + (x → α))
                                                    } yield rt                    
          }
  \end{lstlisting}
\caption{The generic interpreter,
  shared by the unstaged/staged + concrete/abstract interpreter.}
\label{fig:shared_int}
\end{figure}

\paragraph{The Interpreter} Now we construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as reader and state.  Note that the
interpreter is written in open-recursive style -- it can not refer to
itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method
@close@ that lifts lambda terms to closures and method @ap_clo@ that
applies functions also takes an extra @ev@, because evaluation may
happen inside of them.
To close the open-recursion, we use a fixed-point opertor @fix@.
For concrete-interpretation instantiation, it works like the Y
combinator; for abstract-interpretation instantiation, it instruments
the interpreter by memorizing its input and output, which will ensure
the termination of abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also
an abstract type.

%\vspace{-1em}

### 2019-08-14 paper/50_stageabs.tex
revision

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistent, i.e., the generated program does not change the observational
    behaviours of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal form that preserves the
    execution order within a stage \todo{cite WF16}. However, in general, it is
    possible to subvert these guarantees by using unsafe features, such as
    casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularlly.
  \item We construct the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structure we used here are simply
    a black-box that wraps the Scala's library data structures (in the next
    stage).
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

>>>>

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistent, i.e., the generated program does not change the observational
    behaviours of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularlly.
  \item We construct the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structure we used here are simply
    a black-box that wraps the Scala's library data structures (in the next
    stage).
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

### 2019-08-14 paper/80_related.tex
revision

Splitting an analysis into multiple stages is also studied for analysis other
than abstract-interpretation-based control-flow analysis, though the formulation may
very different. \citet{DBLP:conf/cgo/HardekopfL11} apply staging to a
flow-sensitive pointer analysis. The first stage is to analyze the program code
to obtain a sparse representation, and then the second stage conducts the
flow-sensitive analysis by refining the first one. Another large class of static
analysis are implemented by generating constraints firstly and then let a solver
discharge these proof obligations or find the fixed-point. Examples include SAT-
and SMT-based program analysis\todo{cite}, Datalog-based points-to
analysis\todo{cite}, and \todo{}. Similar to the idea in this paper, such
analyses can be splitted to staged.
One of the future work is to explore that whether these analyses can be
implemented and optimized from metaprogramming. Souffle, one of the
state-of-the-art Datalog engine that is commonly used in points-to analysis,
uses the idea of Futamura projection for efficent implemetation. But it only
specializes Datalog rules to the running engine, instead of a full pipeline from
program AST to generating constraints and solving constraints.

Recent work shows that the numerical abstract domains can be also specialized
w.r.t. the static program structure. For example, decomposing polyhedras
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to smaller ones
depending on the variables involved in a statement can significantly improve the
running time and space of abstract transformers.

@@

\paragraph{Two-level Semantics} The idea of reinterpreting the semantics as
abstract interpretation can be traced to Nielson's two-level semantics
\cite{NIELSON1989117}; using two-level semantics for code generation was also
explored by \citet{NIELSON198859}. \citet{Sergey:2013:MAI:2491956.2491979}'s
work of monadic abstract interpreters is also closely related to the two-level
semantics: the use of a generic interface with monads and then reinterpret it by
different semantics is already two-level. Instead of focusing on semantics, this
paper shows how a staged analyzer can be built and used to increase efficieny of
static analysis, we augment the monadic abstract interpreter from another
dimension, using code generation to produce efficient low-level code.

>>>>

Splitting an analysis into multiple stages is also studied for
analysis other than abstract-interpretation-based control-flow
analysis, though the formulation may very
different. \citet{DBLP:conf/cgo/HardekopfL11} apply staging to a
flow-sensitive pointer analysis. The first stage is to analyze the
program code to obtain a sparse representation, and then the second
stage conducts the flow-sensitive analysis by refining the first
one. Another large body of static analyses are implemented by
generating constraints firstly and then let a solver discharge these
proof obligations or find the fixed-point. Notable examples include
SAT- and SMT-based program analysis
\cite{Gulwani:2008:PAC:1375581.1375616} and Datalog-based points-to
analysis \cite{Smaragdakis:2015:PA:2802194.2802195}.
Similar to the idea in this paper, such analyses can be splitted to
staged.  One of the future work is to explore that whether these
analyses can be implemented and optimized from
metaprogramming. Souffl{\'e}, one of the state-of-the-art Datalog
engine that is commonly used in points-to analysis
\cite{Antoniadis:2017:PDS:3088515.3088522}, uses the idea of Futamura
projection for efficent implemetation \cite{10.1007/978-3-319-41540-6_23}.
But it only specializes Datalog rules to the running engine, instead
of a full pipeline from program AST to generating constraints and
solving constraints.

Recent work shows that the numerical abstract domains can be
specialized w.r.t. the static program structure, although not in the
way of using staging. For example, decomposing polyhedras
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143} to
smaller ones that depend on the variables involved in a statement can
significantly improve the running time and space of abstract
transformers.

@@

\paragraph{Two-level Semantics} The idea of reinterpreting the semantics as
abstract interpretation can be traced to Nielson's two-level semantics
\cite{NIELSON1989117}; using two-level semantics for code generation
was also explored by \citet{NIELSON198859}.
\citet{Sergey:2013:MAI:2491956.2491979}'s work of monadic abstract
interpreters is also closely related to the two-level semantics: the
use of a generic interface with monads and then reinterpret it by
different semantics is already two-level. Instead of focusing on
semantics, this paper shows how a staged analyzer can be built and
used to increase efficieny of static analysis, we augment the monadic
abstract interpreter from another dimension, using code generation to
produce efficient low-level code. Hence the presented work can be
considered as an two-dimension application of the two-level semantics.
